Convex Surrogates via Polyhedral Losses

### 2019-02-01 cvx-fin.tex
pass through proof of abstain loss

\subsection{Applications: Abstain Loss and 0-1 Loss \raf{or something}}

\begin{theorem}[\bo{cite}]
  The mode with $|\Y|=n$ is not $(n-2)$-embeddable.
\end{theorem}
\begin{proof}
  Call the outcomes $\Y = \{1,\dots,n\}$.
  We will pick the level set for the report $r=1$ and show it has no $d$-representation.
  The level set can be represented as $\{p : p_1 \geq p_i ~ (\forall i \neq 1)\}$, or the set of $p$ such that $Bp \geq 0$ with
  \[ B = \begin{pmatrix}
       1  & -1 &  0 & \cdots & 0  \\
       1  &  0 & -1 & \cdots & 0  \\
          &    &    & \vdots &    \\
       1  &  0 &  0 & \cdots & -1
     \end{pmatrix} .
  \]
  Here $B$ has $n-1$ rows, one for each $i \neq 1$.

  A $d$-representation is therefore a $V$ with $n-1$ rows, each a unit vector $v_i$ in $\reals^d$.
  Suppose $d \leq n-2$, i.e. there are at least $d+1$ rows.
  Now consider the halfspace representation given by the second column of $B$, namely $T = \{ w : Vw \leq (-1, 0, \dots, 0)^{\intercal}\}$.
  The first $d+1$ rows are linearly dependent in $\reals^d$.
  Therefore, we can write the $d+1$st row as $v_{d+1} = \alpha_1 v_1 + \cdots + \alpha_d v_d$.
  So if $w \in T$, then $\langle v_1,w \rangle \leq -1$ and $\langle v_i,w \rangle \leq 0$ for all $i=2,\dots,d$.
  Taking the weighted combination, we obtain $\langle v_{d+1}, w \rangle \leq -\alpha_1$.
  If $\alpha_1$ is positive, this implies there is slack in the $d+1$st constraint, so the halfspace representation cannot be tight.
  If $\alpha_1$ is nonpositive, then we pick some other pair of vectors $i \neq j$ rather than $1,d+1$, such that $v_j = \sum_{i'\neq j} \alpha_{i'} v_{i'}$ with $\alpha_i > 0$.
  \bo{I claim there must exist two such vectors in this collection ... need to prove it though!}
  Choosing the $i$th column of $B$ as a constraint gives the contradiction as above.
\end{proof}
To recall what is happening inside this proof: we chose the first level set, which gave us a matrix $B$.
The first column was all-$1$, which corresponds to a constraint on the subgradient set of the loss on outcome $1$ for report $1$, which is not an interesting constraint.
The other constraints gave a contradiction if $d$ is too small.
\bo{More to say?}

\begin{theorem}
  Abstain with $\alpha=1/2$ and $|\Y|=5$ is not $2$-embeddable.
\end{theorem}
\begin{proof}
\jessie{Taking a crack at it.}
By Corollary~\ref{cor:d-embed-implies-d-construction}, we want to show there is no $2$-construction for the abstain cell $B^{(abs)} := \ones - 2D$, where $\ones$ is the $5\times 5$ matrix of all entries being $1$, and $D$ being the matrix with $1$ on the diagonal and $0$ elsewhere.

We then want to show there is no $V\in \reals^{5\times 2}$ so that for all $i \in \{1, 2, \ldots, 5\}$ that $(V, B_i)$ is a tight H-rep, following from Definition~\ref{def:d-construction}.
Therefore, we show that there is no $V \in \reals^{5 \times 2}$ such that $S_i := \{w : Vw \leq B_i \}$ is nonempty and for $j = 1, \ldots, 5$, there is a $x \in \reals^2$ such that $\langle v_j, x \rangle = B_{ij}$.

\jessie{This is from the PDF notes...}
In fact, in order for the H-reps to be tight, we must observe that for all $i, j$, the $\max_{x \in S_i} \langle v_j, x \rangle = -1$ if $i = j$ and $1$ otherwise.
For all $i,j$, we denote $x_{ij}$ so that $\langle v_i, x_{ij} \rangle = \langle v_j, x_{ij} \rangle = 1$.
Numbering the normals clockwise (modulo 5), then for all $i$, we need $\langle v_i, x_{i+1, i+2} \rangle \leq -1$ and $\langle v_i, x_{i-1, i-2} \rangle \leq -1$.

Focusing on $v_1$, we know that $\langle v_1, x_{23} \rangle \leq -1$ and $\langle v_1, x_{45} \rangle \leq -1$ if and only if $\theta_{13} = \theta_{12} + \theta_{23} \geq 180^o$  and $\theta_{41} = \theta_{51} + \theta_{45} \geq 180^o$.
These statements are only jointly true if $\theta_{13} = \theta_{41} = 180^o$.
However, the only way to reach this conclusion is if $\theta_{34} = 0^o$, and as the vectors are all unit vectors, we conclude $v_3 = v_4$.
We can see that this scenario contradicts the tightness of the H-reps $(V, B_3)$ and $(V, B_4)$.
As $S_3 = \{w : Vw \leq (1,1,-1,1,1)^T\}$, if $v_3 = v_4$, then the polytope $S_3$ is a line in $\reals^2$ and the constraint $\langle v_1, x \rangle = 1$ is not true for any $x \in S_3$.
Therefore, there is some $i$ such that $(V,S_i)$ is not a tight H-representation.
Thus, the cell $B^{(abs)}$ does not have a $2$-construction, and the property is not $2$-embeddable.

%Similarly, we can see that $\langle v_2, x_{34} \leq -1$ and $\langle v_2, x_{15} \rangle \leq -1$ if and only if $v_4 = -v_2$.
%Therefore, we know that $\theta_{24} = 180^o = \theta_{23} + \theta_{34} = \theta_{12} + \theta{51} + \theta_{45}$
%
%Now, in order for $\langle v_5, x_{12} \leq -1$ and $\langle v_2, x_{34} \rangle \leq -1$ to hold, 
%
%Therefore, for all $v_i$, there must be some $v_j = -v_i$ in $V$.
%
%Consider $V = (v_1, \ldots, v_5)^T$ and, without loss of generality, let $v_1 = (0,1)$.
%
%
%We assert that $v_2 := -v_1$ must be in the set of normals.
%This can be seen since, for the normals $v_1$ and $v_2$, then angle $\theta_{12}$ between the two normals must be $180^o$.
%Since we have $5$ normals, the angle between normals $v_1$ and $v_2$ can be written $\theta{12} = \theta{1i} + \theta_{i2}$.
%(One direction of the angles must have some $v_i$ between $v_1$ and $v_2$.)
%If $\theta_{1i} \leq 90^o$, then $\theta_{1i} + \theta_{i2} \geq 180^o$ in order for there to be a $w$ such that $\langle v_2, w \rangle = 1$.
%
%
%\jessie{Left off here before the reading group}
%In order for the H-rep to be tight, then for all $v_j$ for $j > 1$, $\langle v_j, w \rangle = 1$ and $w = (x, -1)$.
%Observe that a line (here, a facet of the polytope) in $\reals^2$ cannot be nonredundantly bounded by $3$ points.
%This suggests that, in order for $(V,B_1)$ to be tight, at least two facets must both intersect $\langle v_1, w \rangle$ at the same point.
%However, since each $v_j$ must be a unit normal and the facets must intersect at the same point $w$, $\langle v_j, w \rangle = \langle v_k, w \rangle$ for $j \neq k$ if and only if $v_j = v_k$.
%However, if $v_j = v_k$, then the representations $(V, B_j)$ and $(V, B_k)$ will not be tight. 
%This can be seen as there will be no $w \in \{ w: \langle v_j, w \rangle = 1 \}$ that is also on the face $\{w : \langle v_j, w \rangle =-1 \} = \{w : \langle v_k, w \rangle = -1\}$, and vice versa.
%Therefore, there is no set of normals $V \in \reals^{5\times 2}$ so that $\{w : Vw = B_i\}$ is tight for all $v_j$ and $B_i$.
%Therefore, since there is no $2$-construction of abstain(1/2) on $5$ outcomes, we conclude it is not $2$-embeddable.
%\jessie{Weird to think of it this way after drawing so many pictures, but this might be more elegant if it's close.}
\end{proof}

>>>>

\subsection{Examples}

\begin{theorem}[\bo{cite}]
  \bo{Still not proven}
  The mode with $|\Y|=n$ is not $(n-2)$-embeddable.
\end{theorem}
\begin{proof}
  Call the outcomes $\Y = \{1,\dots,n\}$.
  We will pick the level set for the report $r=1$ and show it has no $d$-construction.
  The level set can be represented as $\{p : p_1 \geq p_i ~ (\forall i \neq 1)\}$, or the set of $p$ such that $Bp \geq 0$ with
  \[ B = \begin{pmatrix}
       1  & -1 &  0 & \cdots & 0  \\
       1  &  0 & -1 & \cdots & 0  \\
          &    &    & \vdots &    \\
       1  &  0 &  0 & \cdots & -1
     \end{pmatrix} .
  \]
  Here $B$ has $n-1$ rows, one for each $i \neq 1$.

  A $d$-construction is therefore a $V$ with $n-1$ rows, each a unit vector $v_i$ in $\reals^d$.
  Suppose $d \leq n-2$, i.e. there are at least $d+1$ rows.
  Now consider the halfspace representation given by the second column of $B$, namely $T = \{ w : Vw \leq (-1, 0, \dots, 0)^{\intercal}\}$.
  The first $d+1$ rows are linearly dependent in $\reals^d$.
  Therefore, we can write the $d+1$st row as $v_{d+1} = \alpha_1 v_1 + \cdots + \alpha_d v_d$.
  So if $w \in T$, then $\langle v_1,w \rangle \leq -1$ and $\langle v_i,w \rangle \leq 0$ for all $i=2,\dots,d$.
  Taking the weighted combination, we obtain $\langle v_{d+1}, w \rangle \leq -\alpha_1$.
  If $\alpha_1$ is positive, this implies there is slack in the $d+1$st constraint, so the halfspace representation cannot be tight.
  If $\alpha_1$ is nonpositive, then we pick some other pair of vectors $i \neq j$ rather than $1,d+1$, such that $v_j = \sum_{i'\neq j} \alpha_{i'} v_{i'}$ with $\alpha_i > 0$.
  \bo{I claim there must exist two such vectors in this collection ... need to prove it though!}
  Choosing the $i$th column of $B$ as a constraint gives the contradiction as above.
\end{proof}
To recall what is happening inside this proof: we chose the first level set, which gave us a matrix $B$.
The first column was all-$1$, which corresponds to a constraint on the subgradient set of the loss on outcome $1$ for report $1$, which is not an interesting constraint.
The other constraints gave a contradiction if $d$ is too small.
\bo{More to say?}

\begin{theorem}
  Abstain with $\alpha=1/2$ and $|\Y|=5$ is not $2$-embeddable.
\end{theorem}
\begin{proof}
By Corollary~\ref{cor:d-embed-implies-d-construction}, we can show there is no $2$-construction for one of the level sets.
We choose the abstain cell, which is $\{p : \max_y p_y \leq 0.5\}$.
Because this equals $\{p : (\forall y) 2p_y \geq \sum_{y'} p_{y'}$\}, we can write the level set as $\{p : Bp \geq 0\}$ for
  \[ B = \begin{pmatrix}
       -1 &  1 &  1 & \cdots & 1  \\
       1  & -1 &  1 & \cdots & 1  \\
          &    &    & \vdots &    \\
       1  &  1 &  1 & \cdots & -1
     \end{pmatrix} .
  \]
In other words, $B := \ones - 2I$, where $\ones$ is the all-ones matrix and $I$ is the identity.
In our example, $|\Y| = 5$, so $B$ has width $5$ because there are $5$ outcomes and height $5$ because there are $5$ constraints (one per outcome).

We then want to show there is no $V\in \reals^{5\times 2}$ so that for all $i \in \{1, 2, \ldots, 5\}$, $(V, B_i)$ is a tight H-rep, following Definition~\ref{def:d-construction}.
%Therefore, we show that there is no $V \in \reals^{5 \times 2}$ such that $S_i := \{w : Vw \leq B_i \}$ is nonempty and for $j = 1, \ldots, 5$, there is a $x \in \reals^2$ such that $\langle v_j, x \rangle = B_{ij}$.

Geometrically, a $2$-construction consists of $5$ unit vectors in the plane. (There are $5$ because there are $5$ rows of $B$.)
Once we place these vectors (picture e.g. the regular pentagon), for each column $B_y$ of $B$, we create a polytope $T_y = \{w : Vw \leq B_y\}$.
Here each unit vector $v_i$ gives us the halfspace $v_i \cdot w \leq B_{iy}$.
For example, if the column $B_y = (1,1,1,1,1)^{\intercal}$, then these constraints are all of the form $v_i \cdot w \leq 1$ and $T_y$ would be the regular pentagon (where the unit vectors are at the midpoints of each side).
In this case, $B_y$ has a single $-1$, so we get the regular pentagon but cut off from one side, leaving a triangle.
We can see in this case that some constraints are strictly redundant; the halfspaces $v_i \cdot w \leq 1$ do not all support this triangle.
Therefore, the H-rep $(V,B_y)$ is not tight.
So the regular pentagon cannot be a $2$-construction of $B$.

\bo{I think a figure would be really great here. One showing the 5 unit vectors, and one showing the constraints we get from one column of $B$ (i.e. the triangle.}

In the remainder of the proof, we extend the above to show that no configuration of $5$ unit vectors produces a tight $2$-construction.

Let $v_1,\dots,v_5$ be any $5$ different unit vectors in $\reals^2$, rows of a matrix $V$.
Let $S_y = \{w : Vw \leq B_y\}$ for each column $B_y$ of $B$, $y\in\{1,\dots,5\}$.
Recall we wish to prove that not all H-reps $(V,B_y)$ are tight, or in other words, there is some strictly loose constraint in one of the definitions of $S_y$.

WLOG, suppose $v_1 = (0,1)$ and they are numbered clockwise.
Let $\theta_{i,j}$ be the clockwise angle between $v_i$ and $v_j$.
The vectors point to the sides of an irregular pentagon $S^* = \{w : Vw \leq \vec{1}\}$.
Let $w_{1,2}$ be the corner of this pentagon between $v_1$ and $v_2$, and so on.
In our case, each $S_y$ is the intersection of $S^*$ with a halfspace, $S_y = \{w \in S^* : v_y \cdot w \leq -1\}$.
So in particular, $(V,B_1)$ is tight only if all three vertices $w_{23},w_{34},w_{45}$ satisfy $v_1 \cdot w \leq -1$.
(If for example $v_1 \cdot w_{23} > -1$, then the constraint $v_2 \cdot w \leq 1$ is not tight, i.e. there is a separation between it and $S_1$.)
This implies $\theta_{13} \geq 180^{\circ}$, but it also implies $\theta_{41} \geq 180^{\circ}$, which is a contradiction (both vectors would have to lie directly opposite $v_1$, but then they would be equal).
\end{proof}

%%%% Previous versions of proofs:
%In fact, in order for the H-reps to be tight, we must observe that for all $i, j$, the $\max_{x \in S_i} \langle v_j, x \rangle = -1$ if $i = j$ and $1$ otherwise.
%For all $i,j$, we denote $x_{ij}$ so that $\langle v_i, x_{ij} \rangle = \langle v_j, x_{ij} \rangle = 1$.
%Numbering the normals clockwise (modulo 5), then for all $i$, we need $\langle v_i, x_{i+1, i+2} \rangle \leq -1$ and $\langle v_i, x_{i-1, i-2} \rangle \leq -1$.
%
%Focusing on $v_1$, we know that $\langle v_1, x_{23} \rangle \leq -1$ and $\langle v_1, x_{45} \rangle \leq -1$ if and only if $\theta_{13} = \theta_{12} + \theta_{23} \geq 180^o$  and $\theta_{41} = \theta_{51} + \theta_{45} \geq 180^o$.
%These statements are only jointly true if $\theta_{13} = \theta_{41} = 180^o$.
%However, the only way to reach this conclusion is if $\theta_{34} = 0^o$, and as the vectors are all unit vectors, we conclude $v_3 = v_4$.
%We can see that this scenario contradicts the tightness of the H-reps $(V, B_3)$ and $(V, B_4)$.
%As $S_3 = \{w : Vw \leq (1,1,-1,1,1)^T\}$, if $v_3 = v_4$, then the polytope $S_3$ is a line in $\reals^2$ and the constraint $\langle v_1, x \rangle = 1$ is not true for any $x \in S_3$.
%Therefore, there is some $i$ such that $(V,S_i)$ is not a tight H-representation.
%Thus, the cell $B^{(abs)}$ does not have a $2$-construction, and the property is not $2$-embeddable.

%Similarly, we can see that $\langle v_2, x_{34} \leq -1$ and $\langle v_2, x_{15} \rangle \leq -1$ if and only if $v_4 = -v_2$.
%Therefore, we know that $\theta_{24} = 180^o = \theta_{23} + \theta_{34} = \theta_{12} + \theta{51} + \theta_{45}$
%
%Now, in order for $\langle v_5, x_{12} \leq -1$ and $\langle v_2, x_{34} \rangle \leq -1$ to hold, 
%
%Therefore, for all $v_i$, there must be some $v_j = -v_i$ in $V$.
%
%Consider $V = (v_1, \ldots, v_5)^T$ and, without loss of generality, let $v_1 = (0,1)$.
%
%
%We assert that $v_2 := -v_1$ must be in the set of normals.
%This can be seen since, for the normals $v_1$ and $v_2$, then angle $\theta_{12}$ between the two normals must be $180^o$.
%Since we have $5$ normals, the angle between normals $v_1$ and $v_2$ can be written $\theta{12} = \theta{1i} + \theta_{i2}$.
%(One direction of the angles must have some $v_i$ between $v_1$ and $v_2$.)
%If $\theta_{1i} \leq 90^o$, then $\theta_{1i} + \theta_{i2} \geq 180^o$ in order for there to be a $w$ such that $\langle v_2, w \rangle = 1$.
%
%
%\jessie{Left off here before the reading group}
%In order for the H-rep to be tight, then for all $v_j$ for $j > 1$, $\langle v_j, w \rangle = 1$ and $w = (x, -1)$.
%Observe that a line (here, a facet of the polytope) in $\reals^2$ cannot be nonredundantly bounded by $3$ points.
%This suggests that, in order for $(V,B_1)$ to be tight, at least two facets must both intersect $\langle v_1, w \rangle$ at the same point.
%However, since each $v_j$ must be a unit normal and the facets must intersect at the same point $w$, $\langle v_j, w \rangle = \langle v_k, w \rangle$ for $j \neq k$ if and only if $v_j = v_k$.
%However, if $v_j = v_k$, then the representations $(V, B_j)$ and $(V, B_k)$ will not be tight. 
%This can be seen as there will be no $w \in \{ w: \langle v_j, w \rangle = 1 \}$ that is also on the face $\{w : \langle v_j, w \rangle =-1 \} = \{w : \langle v_k, w \rangle = -1\}$, and vice versa.
%Therefore, there is no set of normals $V \in \reals^{5\times 2}$ so that $\{w : Vw = B_i\}$ is tight for all $v_j$ and $B_i$.
%Therefore, since there is no $2$-construction of abstain(1/2) on $5$ outcomes, we conclude it is not $2$-embeddable.
%\jessie{Weird to think of it this way after drawing so many pictures, but this might be more elegant if it's close.}

\paragraph{On the matching upper bound.}
In constrast to this lower bound, \bo{cite} give a $2$-embedding of the abstain loss on $|\Y|=4$.
This implies that a $2$-construction for the $4 \times 4$ version of the above $B$ matrix exists.
Indeed, one can take $V$ to be the four unit vectors pointing directly north, south, east, and west; the H-representations are then line segments, e.g. the segment $w_1=-1, w_2 \in [-1,1]$, etc.
These are all tight.
Indeed, as required by our characterization, the subgradient sets of their loss functions at the abstain embedding point span these H-reps (in fact are equal to them).

\bo{I think we should actually construct the losses for the ``rhombus'' and show that it embeds abstain loss. That might be a cool showcase for our understanding.}
